Covariate analysis of UFL-Mayo-ISB RNASeq reprocessed counts (both cerebellum and temporal cortex) with SVA adjustments

### 2016-05-29 code/Rmd/Mayo_RNASeq_SVA_Reprocessed.Rmd
Added markdown to process reprocessed RNASeq data from Mayo brain bank with two-step sva adjustment

---
title: "Covariate analysis of UFL-Mayo-ISB RNASeq reprocessed counts (both cerebellum and temporal cortex)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)
library(knitr)
library(githubr)

@@

knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_Reprocessed.Rmd",
                                 parentId = "syn5570325",
                                 entityName = 'UFL_Mayo_ISB RNASeq Reprocessed Counts')
```

@@

thisFileName <- 'Mayo_RNASeq_Reprocessed.Rmd'

@@

### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation}
# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=NULL, plot=T)

# Initial design matrix with covariates under consideration
DM = getDesignMatrix(COVARIATES, Intercept = F)
DESIGN = getDesignMat2Fact(DM$design,FactorCovariates)
colnames(DESIGN) = substr(colnames(DESIGN),1,50)
```

@@

tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
```
`r tmp`

### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Adding "BrainRegion.Diagnosis" back to the residuals
```{r addVarsBack}
varsToAddBackIn <-  grep("BrainRegion.Diagnosis", colnames(FIT$coefficients), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
```

@@

# Get differnetial expression
DE = lapply(1:12, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
}, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
names(DE) = gsub('BrainRegion.Diagnosis','',colnames(contrast))

@@

tmp = DE %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  group_by(Comparison) %>%
  dplyr::summarize(nDiffExpGenes = length(unique(genes)))
kable(tmp)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB RNASeq Reprocessed Counts', parentId = parentId)
CODE <- synStore(CODE)

>>>>

---
title: "Covariate analysis of UFL-Mayo-ISB RNASeq reprocessed counts (both cerebellum and temporal cortex) with SVA adjustments"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)
library(knitr)
library(githubr)

@@

knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_SVA_Reprocessed.Rmd",
                                 parentId = "syn5570325",
                                 entityName = 'UFL_Mayo_ISB RNASeq Reprocessed Counts (with SVA adjustments)')
```

@@

thisFileName <- 'Mayo_RNASeq_SVA_Reprocessed.Rmd'

@@

### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation}
# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=NULL, plot=T)
```

@@

tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
```
`r tmp`
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Estimate surrogate variables using SVA package
```{r sva}
MODEL0 = getDesignMatrix(COVARIATES[,setdiff(postAdjustCovars, "BrainRegion.Diagnosis"),drop=F],Intercept = F)$design # Get null model by removing variable of interest
MODEL1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design

## Calculate the number of surrogate variables
# Get number of surrogate variables (for details of the algorithm refer http://biorxiv.org/content/biorxiv/suppl/2016/05/09/052209.DC1/052209-1.pdf)
SINGULAR.VAL.DECOM <- svd(RESIDUAL.GENE_EXPRESSION)
NUM.DEG.FREEDOM = dim(RESIDUAL.GENE_EXPRESSION)[2] - dim(MODEL1)[2] + 1
SING.VALS <- SINGULAR.VAL.DECOM$d^2/sum(SINGULAR.VAL.DECOM$d)
SING.VALS = SING.VALS[1:NUM.DEG.FREEDOM]

# Permute and calculate the permuted singular values
PERM.SING.VALS <- foreach (i = 1:100,
         .combine = rbind,
         .packages = c('limma','CovariateAnalysis'),
         .export = c('EXPR','MODEL1')) %dopar% {

           # Permute expression counts
           PERM.COUNTS <- t(apply(EXPR$filteredExprMatrix, 1, sample, replace = FALSE))

           # Estimate voom weights
           VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)

           # Fit linear model using new weights and new design
           VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)

           # Residuals after normalisation
           RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)

           PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
           PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
        }
num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])

num.sv2 = num.sv(EXPR$filteredExprMatrix$COUNT, MODEL1, method = 'be', seed = 123456)
NUM.SV = min(num.sv1, num.sv2)

SURR.VAR = svaseq(EXPR$filteredExprMatrix$counts, MODEL1, MODEL0, n.sv = NUM.SV, method = "two-step", B = 30)$sv
colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])

# Estimate voom weights
VOOM.ADJUSTED.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=cbind(DM1$design, SURR.VAR), plot=F)

# Re-estimate voom weights
VOOM.ADJUSTED.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix$counts, design=cbind(DM1$design, SURR.VAR), plot=F, 
                                     block = COVARIATES$Donor_ID, correlation = correlation$cor)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
```

### Residual calculation
Adding "BrainRegion.Diagnosis" back to the residuals
```{r addVarsBack}
varsToAddBackIn <-  grep("BrainRegion.Diagnosis", colnames(FIT$coefficients), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
```

@@

# Get differnetial expression
DE = lapply(1:12, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%
    rownameToFirstColumn('Gene.ID')
}, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
names(DE) = gsub('BrainRegion.Diagnosis','',colnames(contrast))

@@

tmp = DE %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  group_by(Comparison) %>%
  dplyr::summarize(nDiffExpGenes = length(unique(Gene.ID)))
kable(tmp)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB RNASeq Reprocessed Counts (with SVA adjustments)', parentId = parentId)
CODE <- synStore(CODE)
